Specific taxa differentiate proximal and distal microbiota in the unprepped human colon

### 2017-06-28 submission/manuscript.tex
easy Pat edits complete

Colorectal cancer (CRC) remains a leading cause of death worldwide
despite improved preventative and therapeutic measures. Tumors of the
proximal (right) and distal (left) colon are morphologically and
genetically distinct. Previous work from our group found that microbial
dysbiosis is associated with the development of CRC tumors in studies of
both mice and humans. Analysis of the fecal microbiota from healthy and
CRC patients further revealed different microbial signatures associated
with disease. We extended our observations of the fecal microbiome to
include analysis of the proximal and distal human colon. We used a
two-colonoscope approach on subjects that had not undergone standard
bowel preparation procedure. This technique allowed us to characterize
the native proximal and distal luminal and mucosal microbiome without
prior chemical disruption. 16S rRNA gene sequencing was performed on
proximal and distal mucosal and luminal biopsies and home-collected
stool for 20 healthy individuals. Diversity analysis revealed that each
site contained a diverse community, and that a patient's samples were
more similar to each other than to that of other individuals. Since we
could not differentiate sites along the colon based on community
structure or community membership alone, we employed the Random Forest
machine-learning algorithm to identify key species that distinguish
biogeographical sites. Random Forest classification models were built
using taxa abundance and sample location and revealed distinct
populations that were found in each location. \emph{Peptoniphilus,
Anaerococcus, Enterobacteraceae, Pseudomonas} and \emph{Actinomyces}
were most likely to be found in mucosal samples versus luminal samples
(AUC = 0.925). The classification model performed well (AUC = 0.912)
when classifying mucosal samples into proximal or distal sides, but
separating luminal samples from each side proved more challenging (AUC =
0.755). The left mucosa was found to have high populations of
\emph{Finegoldia, Murdochiella} and \emph{Porphyromonas}. Proximal and
distal luminal samples were comprised of many of the same taxa, likely
reflecting the fact that stool moves along the colon from the proximal
to distal end. Finally, comparison of all samples to fecal samples taken
at exit uncovered that the feces were most similar to samples taken from
the left lumen, again reflecting the anatomical structure of the colon.
By sampling the unprepped human colon, our results have identified
distinct bacterial populations native to the proximal and distal sides.
Further investigation of these bacteria may elucidate if and how these
groups contribute to differential oncogenesis processes on the
respective sides of the colon.

@@

Colorectal cancer (CRC) is the second-leading cause of cancer-related
deaths in the United States. CRC tumors vary in structure, size,
morbidity and symptomology depending upon their geographical location in
the colon. CRC has recently been described to fall into four molecular
subtypes ((1)). Of these subtypes, tumors that arise on the left side of
the colon are infiltrating lesions that present with painful symptoms.
In contrast, 47\% of CRCs are caused by right-sided colon tumors that
are bulky and project into the lumen, often remaining asymptomatic until
advanced carcinogenesis ((2)). Due to the absence of symptoms,
right-sided tumors have a significantly higher mortality rate ((2)). The
left and right sides of the colon differ in the amount of inflammation
present and the genomic instability of precancerous cells, respectively,
as well as oxygen, pH and the presence of antimicrobial peptides ((3),
(4), (5)). Microenvironments differ not only longitutinally along the
colon, but latitudinally from the epithelium to mucosa to intestinal
lumen, offering several sites for different microbial communities to
flourish. Given this varied physiology of the proximal-distal axis of
the colon, symbiotic microbes and their metabolites likely vary between
sites.

Several recent findings have shown that development and progression of
CRC can be attributed to specific molecular events as a result of
interactions between the gut microbiota and human host ((3)). Our group
and others have found that the stool microbiome of patients with CRC is
distinct from that of healthy people ((6)). Further studies manipulating
the gut microbiome using antibiotics or other chemoagents in mice has
shown that dysbiosis preceded and accelerated the development of CRC
tumors ((7)). Comparison of the specific bacteria present on CRC tumors
with those found on nearby healthy tissue has also identified specific
bacterial species that are tumor-associated ((8)). These species include
the oral pathogens \emph{Fusobacterium nucleatum} and
\emph{Porphyromonas asacharolytica}. Interestingly, these periodontal
pathogens have been highly predictive of whether a patient had CRC
tumors or not in our classification studies ((9)). Because these studies
were performed in mice or examined only shed human stool, they were
unable to analyze paried samples from the proximal and distal sides of
the colon. Similarly, comparisons of on- or off-tumor bacteria rarely
have matched tissue from the other side of the colon from the same
patient, limiting what conclusions can be drawn about the colonic
microbiome overall, let alone at that specific site. Due to these
limitations, the contribution of the gut microbiota to these subtypes is
largely undefined. Characterizing these communities could provide needed
insight into CRC etiology, including how the disruption of the healthy
community could promote the initiation or proliferation of the distinct
left and right CRC tumors.

@@

Proximal and distal lumenal and mucosal samples were collected from 20
healthy humans that had not undergone bowel preparation(Figure 1).
Participants also collected stool at home one week prior to the
procedure. To characterize the bacterial communities present at these
sites, 16S rRNA gene sequencing was performed on extracted DNA from each
sample. The relative abundance of each phyla for each site is shown in
Figure 2A. This data combines the samples from all participants in the
study. Each site is primarily dominated by \emph{Firmicutes} and
\emph{Bacteriodetes}, consistent with known variability in human
microbiome research (cite). Likewise, samples had varying levels of
diversity at each site, irrespective of the individual (Figure 2B). That
is, while some individuals had a more diverse right mucosa, some had a
more diverse left mucosa. Therefore we could not identify a clear
pattern of changes in microbial diversity along the gut axis.

To compare similarity between sides (left or right) or sites (lumen or
mucosa), we calculated theta YC distances from OTU abundances and
compared distances for all individuals. Again, across all patient
samples we observed a range of theta YC distances when comparing sample
locations (Figure 3A) and again those ranges did not follow a clear
pattern on an individual basis. However, when comparing median distances
between the right lumen and mucosa, the left versus right lumen, the
left versus right mucosa and the left lumen and mucosa, we found that
the right lumen and mucosa were most similar to each other than the
other samples (P \textless{} 0.005, Wilcoxon, BH adjustment).

Next, we calculated thetaYC distances to examine how each sample
compared to the home-collected exit stool. Amidst variability between
patients, we did identify significantly smaller thetaYc distance between
the left lumenal sample and the exit stool (Figure 3B, P \textless{}
0.05, Wilcoxon, BH adjustment). Furthermore, there was an even larger
difference in the comparisons of the left mucosa to the exit stool,
indicating that the mucosa is different from the stool as compared to
lumen (P \textless{} 0.0005, Wilcoxon, BH adjustment). To determine what
factors may be driving the differences seen among the samples, we
compared thetaYC distances between samples from all subjects
(interpersonal) versus samples from within one subject (intrapersonal).
We found that samples from one individual were overall much more similar
to each other than to other study subjects (Figure 3C), consistent with
previous human microbiome studies that have sampled multiple sites of
the human colon (\textbf{???}, (12), (13)). Thus interpersonal variation
between subjects drives the differences between samples more than sample
site or location. Overall, the results of the similarity analysis
suggest that the contents of the left lumen are most representative of
stool at exit, and the microbes remaining on the mucosa are much
different.

@@

To identify OTUs that were distinct at each biogeographical site, we
constructed several Random Forest models built on OTU abundances. We
built the first model to classify the lumen versus mucosal samples for
the right and left sides, independently (Figure 4A). The constructed
model used ((Xopt)) features for the right and ((Xopt)) for the left.
The models performed well when classifying these samples (0.8 and 1.0,
respectively). The OTUs that were most predictive of each site are
identified by their greatest mean decrease in accuracy when removed from
the model. For distinguishing the right lumen and mucosa, OTUs from the
\emph{Bacteriodes}, \emph{Actinomyces}, \emph{Psuedomonas} and two OTUs
from the \emph{Enterobacteraceae} genera were differentially abundant
(Figure 4B). The model classifying the left lumen and mucosa identified
OTUs from \emph{Turicibacter}, \emph{Finegoldia}, \emph{Peptoniphilus}
and two OTUs from the \emph{Anaerococcus} genera that could distinguish
lumen from mucosal samples (Figure 4C). These results indicate that
there are fine differences between the different sites of the colon, and
that these can be traced down to specific OTUs on each side.

Next, we built a model to differentiate the left and right microbiome.
The model performed best when distinguishing the right versus left
mucosa (Figure 5A, AUC = 0.912) compared to the right versus left lumen
(AUC = 0.755). The model was able to explain ((X\%)) of the variance.
OTUs that were differentially abundant between the left and right mucosa
included members of the \emph{Porphyromonas}, \emph{Murdochiella},
\emph{Finegoldia}, \emph{Anaerococcus} and \emph{Peptoniphilus} genera
(Figure 5B). The model was less-effective at classifying the right and
left lumen (AUC = 0.755). However it did identify differentially
abundant OTUs such as three members of the \emph{Bacteroides} genus, a
\emph{Clostridium IV} OTU and an \emph{Oscillibacter} OTU (Figure 5C).
This analysis found that some of the same OTUs that are distinct between
the mucosa and lumen also drive sidedness- such as \emph{Anaerococcus}
and \emph{Finegoldia}.

@@

Given that specific bacterial species have been associated with
colorectal cancer, we probed our sample set for these OTUs. Among our
100 samples, the most frequent sequence associated wtih the
\emph{Fusobacterium} genus was OTU179, which aligns via BLASTn to
\emph{Fusobacterium nucleatum subsp animalis}. It was the only sequence
to align with a \emph{F. nucleatum} strain- the only species of
\emph{Fusobacterium} known to have oncogenic properties and be found on
the surfaces of colorectal cancer tumors. ((14)). The
\emph{Fusobacterium} positive samples were located on the right and left
mucosa and ranged up to 1\% of the sample (Figure 6A). OTU00152 is
\emph{Porphyromonas} and the most frequent sequence in that OTU aligns
to \emph{Porphyromonas asacharolytica}, another bacterium commonly
detected and isolated from colorectal tumors. OTU152 was only detected
on the left mucosa, and in fact was one of the OTUs the classification
model identified as separating left and right sides (Figure 6B). Samples
that were positive for \emph{P. asacharolytica} ranged in abundances
from 0.01\% - 16\%. Thus, cancer-associated OTUs could be found in our
sample set of 20 healthy individuals.

@@

Here we identified bacterial taxa that are specific to the lumen and
mucosa of the right and left human colon from samples collected during
unprepared colonoscopy. We found that all locations contained a range of
phyla and a range of diversity, but that there was a wide variability
between subjects. Pairwise comparisons of each of the sites revealed
that the right mucosa and lumen were most similar to each other.
Further, comparison of colonoscopy-collected samples with samples
collected from stool at home showed that the left lumen is most similar
to stool at exit. Random Forest algorithms built on OTU abundances from
each sample identified microbes that are particular to each location of
the colon. Finally, we were able to detect some bacterial OTUs
associated with colorectal cancer in our healthy patient cohort. Using
unprepped colonoscopy and machine learning, we have identified bacterial
phyla specific to the healthy proximal and distal human colon.

@@

To get around the noisiness created by a diverse set of samples, we
built a Random Forest model to identify microbes specific to each side.
For each comparison we identified top X OTUs that were strongly
predictive of one site or another. Generally, OTUs identified in each
location were consistent with known physiological gradients along the
gut axis ((5)). For instance, the right mucosa harbored
mucosa-associated facultative anaerobes such as \emph{Actinomyces} and
\emph{Enterobacteraceae} and the aerobic \emph{Psuedomonas} consistent
with the highest oxygen regions of the colon. The left mucosa was far
more likely to host strictly anaerobic species such as
\emph{Porphyromonas}, \emph{Anaerococcus}, \emph{Finegoldia} and
\emph{Peptoniphilus}. The model was less effective at classifying the
right and left lumenal contents, probably because the samples are
arguably composed of the same material.

We detected \emph{F. nucleatum} and \emph{P. asacharolytica} in 8 and 5
of our subjects, respectively. These bacteria have been shown to be
predictive of colorectal cancer in humans ((9)) and have oncogenic
properties in cell culture and in mice ((17)). Interestingly, while
\emph{F. nucleatum} was found on both sides of the colon, \emph{P.
asacharolytica} was only detected in the left mucosa. Not much is known
about the distribution of \emph{P. asacharolytica} but given its
documented anaerobic characteristics and asacharolytic metabolism, it
might not be surprising that it resides in the less-oygen-rich and
proteinaceous left mucosa ((4)). In studies examining bacteria on
colorectal cancer tumors, \emph{F. nucleatum} is more commonly detected
on right-sided tumors, and distribution of \emph{F. nucleatum} decreases
along the colon to rectum ((18)). Of the (8) (40\%) individuals positive
for \emph{F. nucleatum} in this study, the bacteria was spread across
the right mucosa, left lumen and left mucosa. The presence of \emph{F.
nucleatum} in a healthy individual is not necessarily linked to the
development of future colorectal cancers. Because of the spatial
distribution of the \emph{F. nucleatum} in our sample set, we cannot
develop a model for the role of \emph{F. nucleatum} in the healthy
colon. Data examining bacterial biofilms on CRC tumors suggests that
\emph{Fusobacteria} species are more commonly found both on proximal
tumors and in biofilms, indicating that it is not only the presence of
the bacteria but the organization of the tumor community that
contributes to \emph{Fusobacterium's} role in tumorigenesis ((8)).
Finally, \emph{Fusobacteria} and \emph{Porphyromonas} species have been
known to not only co-occur on CRC tumors but also to synergistically
promote tumorigenesis in an oral cancer model ((19), (20)). Thus,
further analysis of the distribution and activities of these pathogens
may elucidate a mechanism for development of CRC tumors in the proximal
or distal colon.

@@

Microbiome-based diagnostics are increasingly being explored as
non-invasive tools to survey for the development of colon cancer. Random
Forest models have been used by our group and others to increase the
detection sensitivity of CRC tumors. Indeed, our group found that a
classification model that used microbiome data in combination with Fecal
Immunohistochemical Test (FIT) results could correctly identify both
carcinoma and adenoma lesions from communities of stool at exit and it
performed much better than FIT alone. Further work from our lab has
shown that microbiome profiling of the FIT cartridge contents
sufficiently represented the stool community ((9)). One caveat of the
FIT study was that there was not sufficient information to test if a
classification model could differentiate between proximal and distal CRC
tumors based on exit stool sample alone, but we would hypothesize that
would not be effective. Given that our results showed that the stool
most accurately reflects the community of the left lumen, we likely
cannot use Random Forest of stool samples to diagnose any changes in the
proximal or mucosal communities.

@@

At a baseline visit, subjects were consented and given a home collection
stool kit (Source of kit supplies). At least one week prior to the
scheduled colonoscopy, subjects were to collect whole stool at home and
two swabs of a Fecal Immunohistochemical Test cartridge (Polymedco Inc.)
and ship the samples to the University. Notably, subjects did not
undergo any bowel preparation method prior to sampling. On procedure
day, subjects reported to the Michigan Clinical Research Unit at the
University of Michigan Health System. Patients were consciously sedated
using Fentanyl, Versed and/or Benadryl as appropriate. A flexible
sigmoidoscope was first inserted about 25cm into the colon and endoscopy
brush used to collect lumenal/stool contents. Two lumenal samples were
collected and the contents immediately deposited into RNAlater (source)
and flash-frozen in liquid nitrogen. The brushes were withdrawn and
biopsy forceps were used to collect mucosal biopsies on sections of the
colon that were pink and free of stool matter. Three mucosal biopsies
were collected and flash-frozen in RNAlater. These samples comprised the
distal or left colon samples. The sigmoidoscope was then withdrawn and a
pediatric colonoscope was inserted to reach the ascending colon. Samples
were then collected as in the distal colon and the colonoscope
withdrawn. All samples were stored at -80 C until study completion.

@@

Similarity of microbial community structure between sites of the gut.
ThetaYC distances are shown for interpersonal similarities between two
sites -- each point represents one individual. In (A), comparisons of
the right and left mucosal and lumen are shown. In (B), comparisons of
each site to the exit stool are shown.

@@

Random Forest classifies the mucosa and lumen of each side of the colon.
A) Receiver Operator Characteristic curves are shown for the 10-fold
cross validation of the Random Forest model classifying lumen and
mucosal samples for the left and right sides of the colon. (B) Top five
OTUs that are most important for the classification model for the left
mucosa and lumen (B) and the right mucosa and lumen (C).

@@

Random Forest classifies the left and right sides of the colon. A)
Receiver Operator Characteristic curves are shown for the 10-fold cross
validation of the Random Forest model classifying left lumen versus
right lumen (orange) and left mucosa vs right mucosa (green). (B) Top
five OTUs that are most important for the classification model for the
left and right mucosa (B) and the left and right lumen (C).

>>>>

Colorectal cancer (CRC) remains a leading cause of death worldwide
despite improved preventative and therapeutic measures. Tumors of the
proximal (proximal) and distal (distal) colon are morphologically and
genetically distinct. Previous work from our group found that microbial
dysbiosis is associated with the development of CRC tumors in studies of
both mice and humans. Analysis of the fecal microbiota from healthy and
CRC patients further revealed different microbial signatures associated
with disease. We extended our observations of the fecal microbiome to
include analysis of the proximal and distal human colon. We used a
two-colonoscope approach on subjects that had not undergone standard
bowel preparation procedure. This technique allowed us to characterize
the native proximal and distal luminal and mucosal microbiome without
prior chemical disruption. 16S rRNA gene sequencing was performed on
proximal and distal mucosal and luminal biopsies and home-collected
stool for 20 healthy individuals. Diversity analysis revealed that each
site contained a diverse community, and that a patient's samples were
more similar to each other than to that of other individuals. Since we
could not differentiate sites along the colon based on community
structure or community membership alone, we employed the Random Forest
machine-learning algorithm to identify key species that distinguish
biogeographical sites. Random Forest classification models were built
using taxa abundance and sample location and revealed distinct
populations that were found in each location. \emph{Peptoniphilus,
Anaerococcus, Enterobacteraceae, Pseudomonas} and \emph{Actinomyces}
were most likely to be found in mucosal samples versus luminal samples
(AUC = 0.925). The classification model performed well (AUC = 0.912)
when classifying mucosal samples into proximal or distal sides, but
separating luminal samples from each side proved more challenging (AUC =
0.755). The distal mucosa was found to have high populations of
\emph{Finegoldia, Murdochiella} and \emph{Porphyromonas}. Proximal and
distal luminal samples were comprised of many of the same taxa, likely
reflecting the fact that stool moves along the colon from the proximal
to distal end. Finally, comparison of all samples to fecal samples taken
at exit uncovered that the feces were most similar to samples taken from
the distal lumen, again reflecting the anatomical structure of the
colon. By sampling the unprepped human colon, our results have
identified distinct bacterial populations native to the proximal and
distal sides. Further investigation of these bacteria may elucidate if
and how these groups contribute to differential oncogenesis processes on
the respective sides of the colon.

@@

Colorectal cancer (CRC) is the second-leading cause of cancer-related
deaths in the United States. CRC tumors vary in structure, size,
morbidity and symptomology depending upon their geographical location in
the colon. CRC has recently been described to fall into four molecular
subtypes ((1)). Of these subtypes, tumors that arise on the distal side
of the colon are infiltrating lesions that present with painful
symptoms. In contrast, 47\% of CRCs are caused by proximal-sided colon
tumors that are bulky and project into the lumen, often remaining
asymptomatic until advanced carcinogenesis ((2)). Due to the absence of
symptoms, proximal-sided tumors have a significantly higher mortality
rate ((2)). The distal and proximal sides of the colon differ in the
amount of inflammation present and the genomic instability of
precancerous cells, respectively, as well as oxygen, pH and the presence
of antimicrobial peptides ((3), (4), (5)). Microenvironments differ not
only longitutinally along the colon, but latitudinally from the
epithelium to mucosa to intestinal lumen, offering several sites for
different microbial communities to flourish. Given this varied
physiology of the proximal-distal axis of the colon, symbiotic microbes
and their metabolites likely vary between sites.

Several recent findings have shown that development and progression of
CRC can be attributed to specific molecular events as a result of
interactions between the gut microbiota and human host ((3)). Our group
and others have found that the stool microbiome of patients with CRC is
distinct from that of healthy people ((6)). Further studies manipulating
the gut microbiome using antibiotics or other chemoagents in mice has
shown that dysbiosis preceded and accelerated the development of CRC
tumors ((7)). Comparison of the specific bacteria present on CRC tumors
with those found on nearby healthy tissue has also identified specific
bacterial species that are tumor-associated ((8)). These species include
the oral pathogens \emph{Fusobacterium nucleatum} and
\emph{Porphyromonas asacharolytica}. Interestingly, these periodontal
pathogens have been highly predictive of whether a patient had CRC
tumors or not in our classification studies ((9)). Because these studies
were performed in mice or examined only shed human stool, they were
unable to analyze paried samples from the proximal and distal sides of
the colon. Similarly, comparisons of on- or off-tumor bacteria rarely
have matched tissue from the other side of the colon from the same
patient, limiting what conclusions can be drawn about the colonic
microbiome overall, let alone at that specific site. Due to these
limitations, the contribution of the gut microbiota to these subtypes is
largely undefined. Characterizing these communities could provide needed
insight into CRC etiology, including how the disruption of the healthy
community could promote the initiation or proliferation of the distinct
distal and proximal CRC tumors.

@@

Lumenal and mucosal samples were collected from the proximal and distal
colon of 20 healthy humans that had not undergone bowel preparation
(Figure 1). Participants also collected stool at home one week prior to
the procedure. To characterize the bacterial communities present at
these sites, 16S rRNA gene sequencing was performed on extracted DNA
from each sample. Each site was primarily dominated by \emph{Firmicutes}
and \emph{Bacteriodetes} (Figure 2A), consistent with known variability
in human microbiome research (cite). Likewise, samples had varying
levels of diversity at each site, irrespective of the individual (Figure
2B). For example, the proximal mucosa was more diverse than the distal
for some individuals while the opposite was true for others. Therefore
we could not identify a clear pattern of changes in microbial diversity
along the gut axis.

To compare similarity between sides (proximal or distal) or sites (lumen
or mucosa), we calculated \({\theta}\)YC distances from OTU abundances
and compared these distances for all individuals. Again, across all
patient samples we observed a range of \({\theta}\)YC distances when
comparing sample locations (Figure 3A) and again those ranges did not
follow a clear pattern on an individual basis. However, when comparing
median distances between the proximal lumen and mucosa, the proximal
versus distal lumen, the proximal versus distal mucosa, and the distal
lumen and mucosa, we found that the proximal lumen and mucosa were most
similar to each other than the other samples (P \textless{} 0.005,
Wilcoxon, BH adjustment).

\paragraph{Stool at exit most resembles lumenal samples from the distal
colon}\label{stool-at-exit-most-resembles-lumenal-samples-from-the-distal-colon}

Next, we calculated \({\theta}\)YC distances to examine how each sample
compared to the home-collected exit stool. Amidst variability between
patients, we did identify significantly smaller \({\theta}\)YC distance
between the distal lumenal sample and the exit stool (Figure 3B, P
\textless{} 0.05, Wilcoxon, BH adjustment). Furthermore, there was an
even larger difference in the comparisons of the distal mucosa to the
exit stool, indicating that the mucosa is different from the stool as
compared to lumen (P \textless{} 0.0005, Wilcoxon, BH adjustment). To
determine what factors may be driving the differences seen among the
samples, we compared thetaYC distances between samples from all subjects
(interpersonal) versus samples from within one subject (intrapersonal).
We found that samples from one individual were far more similar to each
other than to other study subjects (Figure 3C), consistent with previous
human microbiome studies that have sampled multiple sites of the human
colon (\textbf{???}, (12), (13)). Thus interpersonal variation between
subjects drives the differences between samples more than sample site or
location. Overall, the results comparing the structure of the
communities suggest that the contents of the distal lumen are most
representative of stool at exit, and the microbes remaining on the
mucosa are much different.

@@

To identify OTUs that were distinct at each biogeographical site, we
constructed several Random Forest models trained using OTU abundances.
We built the first model to classify the lumen versus mucosal samples
for the proximal and distal sides, independently (Figure 4A). The
constructed model used ((Xopt)) features for the proximal and ((Xopt))
for the distal. The models performed well when classifying these samples
(0.8 and 1.0, respectively). The OTUs that were most predictive of each
site are identified by their greatest mean decrease in accuracy when
removed from the model. For distinguishing the proximal lumen and
mucosa, OTUs from the \emph{Bacteriodes}, \emph{Actinomyces},
\emph{Psuedomonas} and two OTUs from the \emph{Enterobacteraceae} genera
were differentially abundant (Figure 4B). The model classifying the
distal lumen and mucosa identified OTUs from \emph{Turicibacter},
\emph{Finegoldia}, \emph{Peptoniphilus} and two OTUs from the
\emph{Anaerococcus} genera that could distinguish lumen from mucosal
samples (Figure 4C). These results indicate that there are fine
differences between the different sites of the colon, and that these can
be traced down to specific OTUs on each side.

Next, we built a model to differentiate the proximal and distal lumenal
samples. The model performed best when distinguishing the proximal
versus distal mucosa (Figure 5A, AUC = 0.912) compared to the proximal
versus distal lumen (AUC = 0.755). These models were able to explain
((X\%)) of the variance, respectively. OTUs that were differentially
abundant between the distal and proximal mucosa included members of the
\emph{Porphyromonas}, \emph{Murdochiella}, \emph{Finegoldia},
\emph{Anaerococcus} and \emph{Peptoniphilus} genera (Figure 5B).
Differentially abundant OTUs of the proximal and distal lumen included
three OTUs of the \emph{Bacteroides} genus, a \emph{Clostridium IV} OTU
and an \emph{Oscillibacter} OTU (Figure 5C). This analysis found that
some of the same OTUs that are distinct between the mucosa and lumen
also helped to differentiate between the two sides- such as
\emph{Anaerococcus} and \emph{Finegoldia}.

@@

Given that specific bacterial species have been associated with
colorectal cancer and IBD, we probed our sample set for these OTUs.
Among our 100 samples, the most frequent sequence associated with the
\emph{Fusobacterium} genus was OTU179, which aligns via BLASTn to
\emph{Fusobacterium nucleatum subsp animalis} (XX\% over full length).
This is the only species of \emph{Fusobacterium} known to have oncogenic
properties and be found on the surfaces of colorectal cancer tumors.
((14)). The \emph{Fusobacterium} positive samples were located in x\% of
the the proximal and X\% of the distal mucosa and represented as much as
1\% of any sample (Figure 6A). OTU152 was similar to the members of the
\emph{Porphyromonas} genus and the most frequent sequence in that OTU
aligned to \emph{Porphyromonas asacharolytica} (X\% over full length),
another bacterium commonly detected and isolated from colorectal tumors.
OTU152 was only detected on the distal mucosa, and in fact was one of
the OTUs the classification model identified as separating distal and
proximal sides (Figure 6B). Among the samples that were positive for
\emph{P. asacharolytica}, the relative abundances for this OTU ranged
from 0.01\% - 16\%. Thus, disease-associated OTUs could be found in our
sample set of 20 healthy individuals.

@@

Here we identified bacterial taxa that are specific to the lumen and
mucosa of the proximal and distal human colon from samples collected
during unprepared colonoscopy. We found that all locations contained a
range of phyla and a range of diversity, but that there was a wide
variability between subjects. Pairwise comparisons of each of the sites
revealed that the proximal mucosa and lumen were most similar to each
other. Further, comparison of colonoscopy-collected samples with samples
collected from stool at home showed that the distal lumen is most
similar to stool at exit. Random Forest algorithms built on OTU
abundances from each sample identified microbes that are particular to
each location of the colon. Finally, we were able to detect some
bacterial OTUs associated with colorectal cancer in our healthy patient
cohort. Using unprepped colonoscopy and machine learning, we have
identified bacterial phyla specific to the healthy proximal and distal
human colon.

@@

To get around the noisiness created by a diverse set of samples, we
built a Random Forest model to identify microbes specific to each side.
For each comparison we identified top X OTUs that were strongly
predictive of one site or another. Generally, OTUs identified in each
location were consistent with known physiological gradients along the
gut axis ((5)). For instance, the proximal mucosa harbored
mucosa-associated facultative anaerobes such as \emph{Actinomyces} and
\emph{Enterobacteraceae} and the aerobic \emph{Psuedomonas} consistent
with the highest oxygen regions of the colon. The distal mucosa was far
more likely to host strictly anaerobic species such as
\emph{Porphyromonas}, \emph{Anaerococcus}, \emph{Finegoldia} and
\emph{Peptoniphilus}. The model was less effective at classifying the
proximal and distal lumenal contents, probably because the samples are
arguably composed of the same material.

We detected \emph{F. nucleatum} and \emph{P. asacharolytica} in 8 and 5
of our subjects, respectively. These bacteria have been shown to be
predictive of colorectal cancer in humans ((9)) and have oncogenic
properties in cell culture and in mice ((17)). Interestingly, while
\emph{F. nucleatum} was found on both sides of the colon, \emph{P.
asacharolytica} was only detected in the distal mucosa. Not much is
known about the distribution of \emph{P. asacharolytica} but given its
documented anaerobic characteristics and asacharolytic metabolism, it
might not be surprising that it resides in the less-oygen-rich and
proteinaceous distal mucosa ((4)). In studies examining bacteria on
colorectal cancer tumors, \emph{F. nucleatum} is more commonly detected
on proximal-sided tumors, and distribution of \emph{F. nucleatum}
decreases along the colon to rectum ((18)). Of the (8) (40\%)
individuals positive for \emph{F. nucleatum} in this study, the bacteria
was spread across the proximal mucosa, distal lumen and distal mucosa.
The presence of \emph{F. nucleatum} in a healthy individual is not
necessarily linked to the development of future colorectal cancers.
Because of the spatial distribution of the \emph{F. nucleatum} in our
sample set, we cannot develop a model for the role of \emph{F.
nucleatum} in the healthy colon. Data examining bacterial biofilms on
CRC tumors suggests that \emph{Fusobacteria} species are more commonly
found both on proximal tumors and in biofilms, indicating that it is not
only the presence of the bacteria but the organization of the tumor
community that contributes to \emph{Fusobacterium's} role in
tumorigenesis ((8)). Finally, \emph{Fusobacteria} and
\emph{Porphyromonas} species have been known to not only co-occur on CRC
tumors but also to synergistically promote tumorigenesis in an oral
cancer model ((19), (20)). Thus, further analysis of the distribution
and activities of these pathogens may elucidate a mechanism for
development of CRC tumors in the proximal or distal colon.

@@

Microbiome-based diagnostics are increasingly being explored as
non-invasive tools to survey for the development of colon cancer. Random
Forest models have been used by our group and others to increase the
detection sensitivity of CRC tumors. Indeed, our group found that a
classification model that used microbiome data in combination with Fecal
Immunohistochemical Test (FIT) results could correctly identify both
carcinoma and adenoma lesions from communities of stool at exit and it
performed much better than FIT alone. Further work from our lab has
shown that microbiome profiling of the FIT cartridge contents
sufficiently represented the stool community ((9)). One caveat of the
FIT study was that there was not sufficient information to test if a
classification model could differentiate between proximal and distal CRC
tumors based on exit stool sample alone, but we would hypothesize that
would not be effective. Given that our results showed that the stool
most accurately reflects the community of the distal lumen, we likely
cannot use Random Forest of stool samples to diagnose any changes in the
proximal or mucosal communities.

@@

At a baseline visit, subjects were consented and given a home collection
stool kit (Source of kit supplies). At least one week prior to the
scheduled colonoscopy, subjects were to collect whole stool at home and
two swabs of a Fecal Immunohistochemical Test cartridge (Polymedco Inc.)
and ship the samples to the University. Notably, subjects did not
undergo any bowel preparation method prior to sampling. On procedure
day, subjects reported to the Michigan Clinical Research Unit at the
University of Michigan Health System. Patients were consciously sedated
using Fentanyl, Versed and/or Benadryl as appropriate. A flexible
sigmoidoscope was first inserted about 25cm into the colon and endoscopy
brush used to collect lumenal/stool contents. Two lumenal samples were
collected and the contents immediately deposited into RNAlater (source)
and flash-frozen in liquid nitrogen. The brushes were withdrawn and
biopsy forceps were used to collect mucosal biopsies on sections of the
colon that were pink and free of stool matter. Three mucosal biopsies
were collected and flash-frozen in RNAlater. These samples comprised the
distal or distal colon samples. The sigmoidoscope was then withdrawn and
a pediatric colonoscope was inserted to reach the ascending colon.
Samples were then collected as in the distal colon and the colonoscope
withdrawn. All samples were stored at -80 C until study completion.

@@

Similarity of microbial community structure between sites of the gut.
ThetaYC distances are shown for interpersonal similarities between two
sites -- each point represents one individual. In (A), comparisons of
the proximal and distal mucosal and lumen are shown. In (B), comparisons
of each site to the exit stool are shown.

@@

Random Forest classifies the mucosa and lumen of each side of the colon.
A) Receiver Operator Characteristic curves are shown for the 10-fold
cross validation of the Random Forest model classifying lumen and
mucosal samples for the distal and proximal sides of the colon. (B) Top
five OTUs that are most important for the classification model for the
distal mucosa and lumen (B) and the proximal mucosa and lumen (C).

@@

Random Forest classifies the distal and proximal sides of the colon. A)
Receiver Operator Characteristic curves are shown for the 10-fold cross
validation of the Random Forest model classifying distal lumen versus
proximal lumen (orange) and distal mucosa vs proximal mucosa (green).
(B) Top five OTUs that are most important for the classification model
for the distal and proximal mucosa (B) and the distal and proximal lumen
(C).
